---
title: "Exploratory Single-Cell RNA-seq Analysis of PBMC Data Using Seurat"
author: "Chitra Karki"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
  pdf_document:
    toc: true
    number_sections: true
    fig_crop: false
  word_document:
    toc: true
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig_crop = FALSE)
```


# Introduction
Single-cell RNA sequencing makes it possible to examine gene activity in individual cells, helping researchers understand how different cell types behave within a mixed population. In this study, we use the pbmc_small dataset from the Seurat package as an example to perform a full single-cell analysis workflow.

The goal is to carry out a clear and reproducible analysis that starts with basic data checks and preparation, followed by data scaling, visualization, and grouping of similar cells. We then assign cell identities to these groups. Beyond this core analysis, we also examine how consistent the groupings are, compare the cells to known reference data, and evaluate gene activity patterns linked to specific biological functions. We further identify genes that differ between groups and explore which biological processes they are involved in. Finally, we compare key immune cell groups,myeloid and lymphoid cells,to understand how their functions differ, using statistical tests on sets of related genes.


# Loading Library

```{r echo=TRUE, message=FALSE, warning=FALSE}
rm(list = ls()) # clearning up the R-environment
library(Seurat) 
library(tidyverse) # data transformation 
library(ggplot2) # visualization
library(patchwork)
library(clustree)# clustering evaluation  
library(SingleR)# annotation  
library(celldex) # AUCell 
library(AUCell) # DGE & enrichment 
library(clusterProfiler)
library(org.Hs.eg.db) # Visualization 
library(ggpubr)


```


# Loading Data
```{r}
data("pbmc_small")
df_pbmc = pbmc_small
```


# Exploration
```{r}
# object type
class(df_pbmc)

# dimension 
dim(df_pbmc)

# structure of the seurat object
str(df_pbmc,max.level = 2)
#str(df_pbmc)
```

The class of the pbmc object is Seurat.
The pbmc data contains 230 rows (genes) and 80 columns (cells).
The Seurat objects (pbmc data set) contains nested lists, data-frames and slots. For clarity the depth of 2 is used.



```{R}
# meta data
head(df_pbmc@meta.data)

```

For the first cell total of  70 RNA molecules were detected, with 47 unique RNA molecules or genes. These counts are valuable piece of information for quality check.


## Quality Control Plots
```{r}
# gene names
rownames(df_pbmc)

# searching mitochondrial genes starting with MT on gene names
PercentageFeatureSet(df_pbmc,pattern = "^MT-") # mt for mouse 
```
There are no Mitochondrial(MT) gene contents detected. The higher content of MT genes indicates the cell death or the poor health of the cell. 


```{r}
# violin plot for quality check
VlnPlot(pbmc_small,features = c("nCount_RNA", "nFeature_RNA"))
```

There are three cluster in the plot, probably its referencing the "RNA_snn_res.1" column in meta data, containing 3 cluster at resolution 1.

The three point on the cluster 2 are isolated from rest of the members if its clusters. They might need attention and further analysis is needed.

```{r}
# correlation between ncount_RNA and nFeaature_RNA
FeatureScatter(df_pbmc,feature1 = "nCount_RNA", feature2 = "nFeature_RNA") + geom_smooth(method = "lm", colour = "black")

```

The correlation between the features is 0.81, which is strong. Points at the ends of the linear line might need attention, as the ratio of nCount to nFeatrue is either very low or high from rest of the data points.

Filters on the features can be applied for the quality control based on the domain knowledge.


## Normalization and Feature Selection
The log normalization is performed to normalize the variance of counts of the genes in cells. Default parameter were used for normalization.
```{r}
df_pbmc = NormalizeData(df_pbmc,normalization.method = "LogNormalize",scale.factor = 10000)

# df_pbmc@assays$RNA$scale.data # scaled data location inside the seurate object
```

After the normalization, the next step in the workflow is the identification of highly variable features (genes). Highly expressed genes will have significant variance across the cells. 

```{r message=FALSE, warning=FALSE}
df_pbmc = FindVariableFeatures(df_pbmc, selection.method = "vst", nfeatures = 20)

#?FindVariableFeatures
# vst: variance stabilizing transformation (loess; statistical method under the hood)

# checking the selected features
top20 = VariableFeatures(df_pbmc)

# plotting the selected features with to 10-genes labeling 
p1 = VariableFeaturePlot(df_pbmc) # without label
p2 = LabelPoints(plot = p1, points = top20[1:10], repel = TRUE)

# plot
p2


```

In this analysis, higly variralbe 20 genes were picked.

In the variance plot top 10 genes are plot for clarity of the labels. PPBP, IGLL5, and CD1C are top three highly variable genes, and might act as biomarkers or they could be outlier as filteration was skipped on the data during quality check.

## Scaling and PCA
The linear transformation (scaling), which is to provide similar weights to the datapoints, is performed before dimensional reduction methods like PCA. The scaling shifts the mean and variance of the expression level of genes across cells to 0 and 1.  

```{r}
#?ScaleData
all_genes = rownames(df_pbmc)
df_pbmc = ScaleData(df_pbmc,features = all_genes )
# by default features = variable features; 20 genes as selected in above step.
```

The scaled data is used for linear dimensional reduction, PCA for this study. This produces the principle components and their contribution on explaining the variability in the expression.

```{r message=FALSE, warning=FALSE}
df_pbmc = RunPCA(df_pbmc, features = VariableFeatures(df_pbmc))

# Examine and visualize PCA results a few different ways
print(df_pbmc[["pca"]], dims = 1:5, nfeatures = 5) 
```

```{r}
# loadings for 2 principal components
VizDimLoadings(df_pbmc, dims = 1:2, reduction = "pca")
```

The loading plot for PC_1 and PC_2 helps to identify important genes on the principal components and their direction on pca plot: same direction suggests some level of similarity in the gene expression and opposite direction hints their dis-similarity.

```{r message=FALSE, warning=FALSE}

# elbow plot gives the total variance explained by each principal components. 
ElbowPlot(df_pbmc)
```

The elbow plot is useful to determine the number of principal components that captures desired level cumulative variability. PC_1 captures majority of the signal almost 3-sd and sharply decline for other PCs thereafter. Roughly around 10th PC the curve starts to flat or linear. Hence, 10 PCs might be enough to capture most cumulative variability in the gene expression.

```{r}
DimPlot(df_pbmc, reduction = "pca") + NoLegend()
```

The most variability is captured by PC_1, as we see the distinct cluster at the far right of PC1. As compared to PC_1, PC2 is capturing less variability which is evident by the variability of the date points along PC_2. Moreover, there is overlap between green and the blue data points, on the positive side of the PC_2, which suggests more PCs are required for clear separation. On the other hand, there are red points towards the negative side of the PC_2, which suggest the genes in green different to red genes in some respect.

```{r}
DimHeatmap(df_pbmc, dims = 1, cells = 80, balanced = TRUE, fast = FALSE)
# balanced = TRUE indicated the same number of cells are utilized on either side of the PC_1 (positive and negative)
```

The heat map shows the level of gene expression across the cells. The bottom left has the yellow patch, so these genes (CA2, TUBB1, PPBP, PF4, SDRP,etc.) have high level of expression across the cells, which shows similarity in them. 

Similarly, the top right corner of the heat map shows patches of mid range expression of genes. This could be a distant cluster.

More PCs heat maps can be generated by adjusting dims argument in the `DimHeatmap` function, for instance `dims = 1:10` will give us 10 heat maps. Elbow plot might come handy in this case to determine the number of PCs to utilize.

## Clustering and tSNE/UMAP

Utilizing 10 PCs from Elbow plot, a Shared Nearest Neighbor (SNN) graph is constructed using `FindNeighbors` function. This calculates the closeness of the cells using the local neighbors in the reduced dimension space with out distorting the underlying structure. The Louvian algorithm was then applied with the function `Findcluster` to identify the highly connected clusters of the cells. In order to have a comparison across resolution, 0 to 1 with step size 0.1 was used.

Resolution of 0.9 wad picked interactively with some hit and trials with the UMAP plot, producing 3 clusters. In the plot, the left panel is of resolution 0.7, which has two cluster but seems like the red points at the top form its own cluster but flagged as cluster 0. In the middle panel (resolution 0.9), the isolated data point on top are flagged as distinct cluster. The left panel is of resolution 1, which stays same as of resolution 0.9.   

```{r message=FALSE, warning=FALSE}
df_pbmc = FindNeighbors(df_pbmc, dims = 1:10)
df_pbmc = FindClusters(df_pbmc, resolution = seq(0.1,1,0.1),verbose = FALSE)
df_pbmc =  RunUMAP(df_pbmc, dims = 1:10)
#df_pbmc = RunTSNE(df_pbmc,dims = 1:10,perplexity=10) # errors with this
#DimPlot(df_pbmc,reduction = "tsne",label = TRUE)

Idents(df_pbmc) <- "RNA_snn_res.0.8"
p08 = DimPlot(df_pbmc, reduction = "umap",label = TRUE) + ggtitle("Resolution:0.8")

Idents(df_pbmc) <- "RNA_snn_res.0.9"
p09 = DimPlot(df_pbmc, reduction = "umap",label = TRUE) + ggtitle("Resolution:0.9")

Idents(df_pbmc) <- "RNA_snn_res.1"
p1 = DimPlot(df_pbmc, reduction = "umap",label = TRUE) + ggtitle("Resolution:1")

p08 + p09 + p1
```

The cluster 2 in the UMAP at 0.9 resolution is isolated from the other two cluster. cluster 0 and cluster 1 are close to each other with no overlap and are connected, suggesting they might share expression similarities or transitioning form one cell stage to other.
 

# Clustree Analysis (Resolution Selection)
```{r,fig.height=12, fig.width=10}
clustree(df_pbmc, prefix = "RNA_snn_res.")
```

The vertical arrows between the resolution 0.9 and 1 indicates the cluster stability, with no further branch outs. This suggests 0.9 is the best resolution. 

# Cell Type Annotation
## SingleR (Reference-Based)
The annotation of the cells are performed with correlation-based annotation tool `SingleR` referencing the Human Primary Cell Atlas library. Interestingly, isolated cluster in PCA/UMA analysis is being annotated as patelets (cluster-2). 

```{r}
# referencing library
ref = HumanPrimaryCellAtlasData()

# annotation prediction by SingleR: Spearman correlation
pred = SingleR(test = as.matrix(df_pbmc@assays$RNA$data), 
                ref = ref, 
                labels = ref$label.main)

# annotation labels
df_pbmc$SingleR_label = pred$labels

# annotation plot
DimPlot(df_pbmc, group.by = "SingleR_label", label = TRUE) +
  ggtitle("Cell Type Annotation")
```


## AUCell (Gene Set-Based)
For cellular identity, `AUCell` is utilized to perform gene set scoring. This rank based method calculated the Area Under the Curve (AUC) to predefined gene clusters. Three gene sets were selected utilizing gold standard markers derived form established immunological frameworks (Abbas et al., 2021; Mabbott et al., 2013),  

* lymphoid (CD3D, CD3E, CD4, CD8A)
* myeloid (LYZ, S100A8, S100A9)
* platelets (PPBP, PF4)

```{r message=FALSE, warning=FALSE}
geneSets = list( lymphoid = c("CD3D", "CD3E", "CD4", "CD8A"), 
                 myeloid = c("LYZ", "S100A8", "S100A9"),
                 patelets = c("PPBP","PF4")) 
exprMatrix = GetAssayData(df_pbmc, layer = "counts")
cells_rankings = AUCell_buildRankings(exprMatrix, plotStats = FALSE)
cells_AUC = AUCell_calcAUC(geneSets, cells_rankings)
auc_matrix = t(getAUC(cells_AUC))
df_pbmc = AddMetaData(df_pbmc, auc_matrix) 
FeaturePlot(df_pbmc, features = c("lymphoid", "myeloid","patelets"))

```

Plots shows, lymphoid, myeloid, and platelets cells are distributed in distinct regions of the UMAP.

## Myeloid vs Lymphoid vs Patelets Asignment
```{r message=FALSE, warning=FALSE}
# specfying the labels 
df_pbmc$celltype =
  case_when(df_pbmc$patelets > df_pbmc$myeloid & df_pbmc$patelets > df_pbmc$lymphoid ~ "patelets",
  df_pbmc$myeloid > df_pbmc$lymphoid ~ "myeloid",
  TRUE ~ "lymphoid"
)
  
#plot
DimPlot(df_pbmc, group.by = "celltype", label = TRUE) + 
  ggtitle("myeloid Vs.lymphoid Vs. patelets") +
  xlab("UMAP 1") + ylab("UMAP 2")
```

# Differential Gene Exploration (DGE)
DEG is a statistical comparison used for identifying genes that are expressed differently in across groups of cells. It measures the fold change (how much higher the expression is in one cluster vs others) and p-values(the statistical confidence of differential gene expression across cells).

## Cluster-Level Marker

```{r message=FALSE, warning=FALSE}
# cluster marker indentification
cluster_markers = FindAllMarkers(df_pbmc, only.pos = TRUE)

# top three genes in the each cluster, act as bio-markers
cluster_markers %>%
  group_by(cluster) %>% 
  arrange(p_val_adj) %>% 
  slice_head(n = 3) 
```

The table displays top three gene in each cluster which are highly expressed and statistically significant in expression across the clusters. 

Moreover, they serve as the markers for their respective cluster, for instance LYZ, TYMP and CSTS are markers for cluster 0 (myeloid). cluster-1 and cluster-2 are lymphoid and patelets,and their markers are (LCK, CD3D,CD7) and (GP9, PF4,GNG11), respectively.

The `avg_log2FC` (fold change) value for LYZ gene in cluster is 4.03, which means it is expressed $2^{4.03}$ (nearly 16) times more strongly in cluster 0 then in all other cells.


## Myeloid vs Lymphoid

```{r message=TRUE, warning=FALSE}
Idents(df_pbmc) = df_pbmc$celltype 
dge_ml <- FindMarkers(df_pbmc, ident.1 = "myeloid", ident.2 = "lymphoid") 
head(dge_ml)
```

The `pct.1` value of the leading gene LYZ is 1 i.e. all the cells in myeloid cluster express this gene. On the other hand, the `pct.2` value 0.455 indicates around $46\%$ of the cells in lymphoid cluster express this gene. S100A8 gene has the highest `avg_log2FC` value, 6.049272, suggesting 65 times stronger expression in lymphoid cluster comparison to rest of the cells.

# Pathway Enrichment Analysis
The functional enrichment analysis of the significant genes were performed across the cluster.
```{r fig.height=12, fig.width=10, message=FALSE, warning=FALSE}

# selection of significant genes <0.05
genes_to_test <- cluster_markers %>%
  filter(p_val_adj < 0.05) %>% # significance level 0.05
  dplyr::select(gene, cluster)

# symbol conversion from human readable gene names to their ID
entrez_df <- bitr(genes_to_test$gene, fromType="SYMBOL", toType="ENTREZID", OrgDb=org.Hs.eg.db)

# merge back to keep cluster assignments
genes_with_clusters <- merge(genes_to_test, entrez_df, by.x="gene", by.y="SYMBOL")

# comparative enrichment
comp_go <- compareCluster(ENTREZID ~ cluster, 
                          data = genes_with_clusters, 
                          fun = "enrichGO", 
                          OrgDb = org.Hs.eg.db, 
                          ont = "BP")

# 3. Visualize
dotplot(comp_go) + ggtitle("Pathway Enrichment Comparison among clusters")

```

From the dot plot following are observed,

* cluster 0 has functionality of  MHC-II mediated antigen presentation
* cluster 1 exhibits cell killing
* cluster 2 exhibits coagulation and hemostasis

Cluster 2 has non-overlapping functions while cluster 0 and 1 share some functionalists.

# Visualization of Significant Genes

```{r message=FALSE, warning=FALSE}
# how many genes to display per cluster
n_genes = 3

# slicing tow n_genes per cluster
top_genes =
  dge_ml %>% 
  rownames_to_column(var = "gene") %>% 
  mutate(cluster = ifelse(avg_log2FC > 0, "myeloid", "lymphoid")) %>% 
  group_by(cluster) %>% 
  arrange(p_val_adj) %>% 
  slice_head(n = n_genes) %>% 
  ungroup() %>% 
  dplyr::select(gene) %>% 
  unlist()
  
#Violin Plot
# statistical comparision  of lymphoid, vs myeloid
comp_clust = list( c("lymphoid", "myeloid") )

#subset with out patelets
df_subset = subset(df_pbmc, idents = c("lymphoid", "myeloid"))

# function to generate plots with statistical test
plot_list = lapply(top_genes, function(gene) {
  VlnPlot(df_subset, features = gene, group.by = "celltype", pt.size = 0) +
    # statistical test for the comparison
    stat_compare_means(comparisons = comp_clust, 
                       method = "wilcox.test", 
                       label = "p.format") +
    # adjust y-axis limit to make space forp-value
    ylim(0, max(FetchData(df_pbmc, vars = gene)) + 2) +
    NoLegend()
})

# final figure
CombinePlots(plots = plot_list, ncol = 3)

```

In the violin plot, the top three genes in lymphoid and myeloid clusters are displayed along with p-values from Wilcoxon rank-sum tests. The statistical confidence is high in their differential expression among the clusters. (43), (16) and (24) represent number of genes in the marker lists that were successfully mapped to the GO database.

# Module Scoring (Myeloid Vs. Lymphoid)
Module scoring identifies what a cell is doing, its functionality or pathway. Here the top enriched functions of myeloid (antigen processing and presentation of exogenous peptide antigen) and lymphoid (killing of cells of another organism ) are extracted for module scoring to test whether the pathway is significantly enriched.

```{r fig.height=12, fig.width=10,message=FALSE, warning=FALSE}
# geting top term genes for myeloid (cluster 0) and lymphoid (cluster 1)
results = comp_go@compareClusterResult

# myeloid pathway:top term:cluster 0
myeloid_term = results %>% 
  filter(cluster == 0) %>% 
  slice(1)

#gene names
myeloid_genes = bitr(strsplit(myeloid_term$geneID, "/")[[1]], 
                      fromType="ENTREZID", toType="SYMBOL", OrgDb=org.Hs.eg.db)$SYMBOL

# lymphoid pathway:top term;cluster 1)
lymphoid_term = results %>% 
  filter(cluster == 1) %>% 
  slice(1)

# gene names
lymphoid_genes = bitr(strsplit(lymphoid_term$geneID, "/")[[1]], 
                       fromType="ENTREZID", toType="SYMBOL", OrgDb=org.Hs.eg.db)$SYMBOL

# module score calculation
df_pbmc = AddModuleScore(df_pbmc, 
                          features = list(myeloid_genes, lymphoid_genes), 
                          name = "PathwayScore", 
                          ctrl = 5)

# rename metadata columns for clarity
colnames(df_pbmc@meta.data)[which(colnames(df_pbmc@meta.data) == "PathwayScore1")] <- "myeloid_Pathway"
colnames(df_pbmc@meta.data)[which(colnames(df_pbmc@meta.data) == "PathwayScore2")] <- "lymphoid_Pathway"

# subset on myeloid vs lymphoid
df_subset <- subset(df_pbmc, idents = c("lymphoid", "myeloid"))

# plot:myeloid pathway
p1 <- VlnPlot(df_subset, features = "myeloid_Pathway", pt.size = 0.5) + 
  stat_compare_means(method = "wilcox.test", label.y = 5, label.x = 1.3) + 
  labs(title = "Antigen Processing (myeloid Path)") +
  ylim(-2, 6) + NoLegend()

# Plot:lymphoid Pathway Activity
p2 <- VlnPlot(df_subset, features = "lymphoid_Pathway", pt.size = 0.5) + 
  stat_compare_means(method = "wilcox.test", label.y = 7, label.x = 1.3) + 
  labs(title = "Cell Killing (lymphoid Path)") +
  ylim(-2, 8) + NoLegend()

# combined plot
p1 + p2

```

The enriched pathway in myeloid and lymphoid cells are statistically significant. The significance of the lymphoid pathway is easy to observe through the violin plots (bottom panel), while for myeloid pathway (top panel), the lymphoid graph may seem tall but myeloid cluster have higher distribution points greater then the lymphoid cluster points.

# Summary
In this study, a complete analysis of single-cell RNA sequencing data was carried out using the pbmc_small dataset. The workflow covered key steps such as data preparation, visualization, grouping of similar cells, producing results, followed by more detailed analyses.

Different methods used to label cell types identifying three main groups of cells: lymphoid, myeloid, and platelet cells. These groups were clearly separated in clusters when visualized, which supported the reliable analysis.

Analysis of gene differences between groups identified strong marker genes that differentiated heterogenisity of the cell type. Further exploration of biological functions showed clear differences between cell types, such as immune signaling roles in myeloid cells and cell-killing activity in lymphoid cells. These functions were confirmed using gene set–based scoring.

Overall, this work presents a clear and reproducible approach to analyzing single-cell RNA-seq data. By combining cell grouping, cell identification, and functional analysis, this study showed how single cell data can be utilized to understand the biological functions of diversed cells and genes.

# References

* Abbas, A. K., Lichtman, A. H., & Pillai, S. (2021). Cellular and Molecular Immunology. 10th Edition. Elsevier Health Sciences.

* Aibar, S., et al. (2017). "SCENIC: single-cell regulatory network inference and clustering." Nature Methods, 14(11), 1083–1086.

* Aran, D., et al. (2019). "Reference-based analysis of single-cell transcriptomes with SingleR." Nature Immunology, 20(2), 163–172.

* Mabbott, N. A., et al. (2013). "An expression atlas of human primary cells: inference of gene function utilizing microarray data." BMC Genomics, 14(1), 632.

* Satija, R., Farrell, J. A., Gennert, D., Schier, A. F., & Regev, A. (2015). "Spatial reconstruction of single-cell gene expression data." Nature Biotechnology, 33(5), 495–502.

* Yu, G., Wang, L., Han, Y., & He, Q. (2012). "clusterProfiler: an R package for comparing biological themes among gene clusters." OMICS: A Journal of Integrative Biology, 16(5), 284–287.



